---
title: "Simulated Randomized Trial Walkthrough"
author: "David McCoy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulated Randomized Trial Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Goal

This article walks through a small randomized trial where the active treatment
improves a binary response endpoint, but the treatment effect is much larger in
participants with `marker_high == 1`. The goal is to show how a trialist should
read the main `rct = TRUE` outputs:

- fold-level selected subgroup rules
- pooled effects in `V`, `V^c`, and the `V - V^c` contrast
- selected-region arm counts and observed outcomes
- selection stability diagnostics

The example is intentionally simple. In a real trial, the endpoint definition,
eligible subgroup variables, analysis role, and tuning parameters should be
written in the protocol or SAP before looking at outcomes.

## Simulate a trial

```{r setup}
library(EffectXshift)
library(sl3)

set.seed(20240624)

n <- 600
marker_high <- rbinom(n, 1, 0.5)
age <- round(rnorm(n, mean = 62, sd = 9))
biomarker <- rnorm(n)
prior_tx <- rbinom(n, 1, plogis(-0.5 + 0.4 * biomarker))

# 1:1 randomized treatment
treatment <- rbinom(n, 1, 0.5)
```

The true response risk is generated so the treatment improves response by about
35 percentage points when `marker_high == 1`, and by about 0 percentage points
otherwise.

```{r outcome}
true_effect <- ifelse(marker_high == 1, 0.35, 0.00)

baseline_response <- plogis(
  -1.2 - 0.02 * (age - 62) + 0.3 * biomarker - 0.2 * prior_tx
)
response_prob <- pmin(pmax(baseline_response + true_effect * treatment, 0.01), 0.99)
response <- rbinom(n, 1, response_prob)

w <- data.frame(
  marker_high = marker_high,
  age = age,
  biomarker = biomarker,
  prior_tx = prior_tx
)
a <- data.frame(treatment = treatment)

aggregate(true_effect ~ marker_high, data = data.frame(marker_high, true_effect), mean)
```

## Run the randomized-trial workflow

For a trial, pass the known allocation probability through `alpha`. Here
`alpha = 0.5` because this is a 1:1 randomized design.

```{r fit}
mu_learner <- list(
  Lrnr_ranger$new(num.trees = 150),
  Lrnr_glm$new()
)

trial_results <- EffectXshift(
  w = w,
  a = a,
  y = response,
  deltas = 0,
  outcome_type = "binary",
  mu_learner = mu_learner,
  n_folds = 2,
  parallel = FALSE,
  seed = 12,
  rct = TRUE,
  rct_type = "ate",
  alpha = 0.5,
  min_obs = 50,
  max_depth = 1,
  pval_thresh = 0.3
)
```

`pval_thresh = 0.3` is intentionally permissive for this small teaching
simulation. In applied work, treat the split rule as a tuning choice and state
it before running the analysis.

## Fold-level rules

```{r kfold}
kfold <- trial_results$`Effect Modification K-Fold Results`
kfold$Rule <- sub("&\\s*$", "", kfold$Rule)
kfold
```

Interpretation:

- `RegionType == "V"` is the discovered higher-effect region in that fold.
- `RegionType == "Vc"` is the held-out complement.
- `Rule` is the selected baseline subgroup rule. Here, `marker_high != 0` means
  the selected region is the `marker_high == 1` subgroup.
- `Effect` is the held-out region treatment-effect estimate on the response-risk
  difference scale.
- `N_in_Region` is the number of validation-fold participants in that region.

The fold-level table is the first place to look. If different folds select
different variables or opposite orientations, the pooled result should be
reported as exploratory and interpreted cautiously.

## Pooled region effects

```{r pooled}
trial_results$`Pooled Region Effects`
```

Interpretation:

- `V` is the treatment effect in the selected subgroup.
- `V^c` is the treatment effect in everyone outside that subgroup.
- `V - V^c` is the effect-modification contrast.

In this simulation, the selected `V` region has a large positive response-risk
difference, while `V^c` is near zero. A positive `V - V^c` contrast indicates
that treatment benefit is larger in the selected subgroup than in its
complement.

## Trial region diagnostics

```{r trial-diagnostics}
trial_results$`Trial Region Diagnostics`
```

These diagnostics are descriptive, not adjusted effect estimates. Use them to
check whether the discovered regions have usable arm counts and whether the
observed outcomes broadly agree with the adjusted results.

Important columns:

- `N_Control`, `N_Treated`: arm counts in the selected region
- `Prop_Treated`: observed treatment fraction; in a randomized trial this should
  be close to the design allocation probability, allowing for random variation
- `Outcome_Mean_Control`, `Outcome_Mean_Treated`: observed endpoint means by arm
- `Observed_Mean_Diff_Treated_Minus_Control`: unadjusted descriptive difference
- `Flag_Small_Arm_N`: quick flag for regions where one arm is too small to
  interpret comfortably

## Selection stability

```{r stability}
diagnose_selection(trial_results)$stability
```

This summarizes whether the same treatment and rule were selected across folds.
Stable fold-level selection supports a more interpretable finding. Unstable
selection does not invalidate the held-out estimation procedure, but it does
mean the discovered subgroup should be framed as exploratory.

## What to write in a report

For this simulated example, an interpretation could read:

> In a 1:1 randomized trial with a binary response endpoint, EffectXshift
> selected the baseline subgroup `marker_high == 1` as the higher-benefit
> region. The pooled held-out treatment effect was larger in `V` than in `V^c`,
> and the `V - V^c` contrast was positive. Arm counts in both selected regions
> were adequate, and fold-level selection was stable across the two folds.

For a real trial, also report whether this was prespecified or exploratory, the
eligible baseline variables, the endpoint time point, censoring or missingness
handling, `alpha`, and all subgroup-search tuning parameters.

